---
title: "In Class Assignment Week 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Don't forget to install the packages in the console!
library(dplyr)
library(nycflights13)
library(ggplot2)
library(tidyr)
```

## Part I: 

1. The data set at rnf6080.dat records hourly rainfall at a certain location in Canada, every day from 1960 to 1980. First, we need to load the data set into R using the command read.table(). Use the help function to learn what arguments this function takes. Once you have the necessary input, load the data set into R and make it a data frame called rain.df.

```{r}
#Load data file
rain.df = as.data.frame(read.table('rnf6080.dat'))
```
2. How many rows and columns does rain.df have? (If there are not 5070 rows and 27 columns, something is wrong; check the previous part to see what might have gone wrong in the previous part.)
```{r}
#Finds number of rows and columns
nrows = nrow(rain.df)
ncols = ncol(rain.df)
nrows
ncols
```

3. What are the names of the columns of rain.df?

```{r}
#Finds names of columns
col_names = names(rain.df)
col_names

```

4. What is the value of row 5, column 7 of rain.df?
```{r}
#Finds value of row 5,7
value_row5_col7 = rain.df[5,7]
value_row5_col7
```

5. Display the second row of rain.df in its entirety.
```{r}
#Prints the second row
print_second_row = rain.df[2,]
print_second_row
```

6. Explain what the command below doesby running it on your data and examining the object. (You may find the display functions head() and tail() useful here.) Is it clear now what the last 24 columns represent?

```
names(rain.df) <- c("year","month","day",seq(0,23))
````

This command changes the labels of the first three columns to "year", "month", and "day" respectively. The last 24 columns represent the hour of the day. 

7. Create a new column in the data frame called daily, which is the sum of the rightmost 24 columns. With this column, create a histogram of the values in this column, which are supposed to be daily rainfall values. What is wrong with this picture?

```{r}

#Adds column
rain.df = 
rain.df %>% 
  mutate(daily = rowSums(rain.df[, 4:27]))
#histogram of daily rainfall
hist(rain.df$daily, main = "Histogram of Daily Rainfall", xlab = "Daily Rainfall Amount", ylab = "Frequency")
```
\   

This histogram has negative values. There can be no negative rainfall amounts. 

8. Create a new data frame rain.df.fixed that takes the original and fixes it for the apparent flaw you have discovered. Having done this, produce a new histogram with the corrected data and explain why this is more reasonable.

```{r}
#Changed all values <0 to NA
rain.df[rain.df < 0] = NA
hist(rain.df$daily, main = "Histogram of Daily Rainfall", xlab = "Daily Rainfall Amount", ylab = "Frequency") 
```
\

This graph is much more reasonable because there are no negative values. 

## Part II:
### Question 1: 
1. Filter all flights on your birthday.
```{r}
#inspect structure of flights dataset
str(flights)

#Part 1:
flights %>% 
  filter(month == 07 & day == 22)
```
### Question 2: 
2. Filter all flights in rows 567 - 589
```{r}
#Part 2:
flights %>%
  filter(row_number() %in% 567:589)
```
### Question 3: 
3. Use arrange() to order the data by year, month and day
```{r}
#Part 3: 
flights %>%
  arrange(year, month, day)
```

#Question 4: 
4. Use desc() to arrange flights by arrival delay.
```{r}
#Part 4:
flights %>% 
  arrange(desc(arr_delay))
```

### Question 5: 
5. Select columns containing data information.
```{r}
#Part 5: 
flights %>% 
  select(1:3)
```

### Question 6: 
6. Select columns including only plane information.
```{r}
#Part 6: 
flights %>%
  select(flight, tailnum, carrier)
```

### Question 7: 
7. Rename tailnum to tail_num
```{r}
#Part 7:
flights %>%
  rename('tail_num' = tailnum)
```

### Question 8: 
8. Create a variable called gain which describes the difference between arrival delap and departure delay.
```{r}
#Part 8: Calculates the total time delayed or gained. 
flights %>%
  mutate(gain = arr_delay - dep_delay)
```

### Question 9:
9. Create a variable called speed which is distance over time in hours.
```{r}
#Part 9: 
flights %>%
  mutate(speed = distance/(air_time/60))
```

### Question 10: 
10. Summarise the average delay for all flights.
```{r}
#Part 10: 
flights %>%
  summarise(mean(arr_delay - dep_delay, na.rm = TRUE))
```
## Overall Questions:
### Question 11:
11. Split the complete dataset into individual planes and then summarise each plane by counting the number of flights (count = n()) and computing the average distance and arrival delay . Then plot this:

```
ggplot(delay, aes(dist, delay)) +
  geom_point(aes(size = count), alpha = 1/2) +
  geom_smooth() +
  scale_size_area()
```

```{r}
delay = 
  flights %>%
  group_by(tailnum) %>%
  summarise(count = n(), dist = mean(distance, na.rm = TRUE), delay = mean(arr_delay, na.rm = TRUE)) 

ggplot(delay, aes(dist, delay)) +
  geom_point(aes(size = count), alpha = 1/2) +
  geom_smooth() +
  scale_size_area()
```

### Question 12: 
12. Find the number of planes and the number of flights that go to each possible destination.
```{r}
#Part 12: 
flights %>% 
  group_by(dest) %>%
  summarise(num_flights = n(), num_planes = n_distinct(tailnum))
```

### Question 13: 
Consider the code Below:

```{r}
filter(
  summarise(
    select(
      group_by(flights, year, month, day),
      arr_delay, dep_delay
    ),
    arr = mean(arr_delay, na.rm = TRUE),
    dep = mean(dep_delay, na.rm = TRUE)
  ),
  arr > 30 | dep > 30
)
```

a. What are some problems with this function? 
b. How easy is it to follow the logic of this? 
c. Rewrite this using piping to make it more understandable. 
d. Does your rewritten command give the same results?

#### Part a:
The problem with this function is that it serves several purposes. The independent functions should be separated for absolute clarity. 

#### Part b: 
Because this one large function does multiple things, the clarity/logic behind it isn't crystal clear. 

#### Part c: 
```{r}
flights %>%
  group_by(year, month, day) %>%
  select(arr_delay, dep_delay) %>%
  summarise(arr = mean(arr_delay, na.rm = TRUE),
            dep = mean(dep_delay, na.rm = TRUE)) %>%
  filter(arr > 30 | dep > 30)
```

### Part d: 
Yes, it gives the same results. 

### Question 14: 
Consider the code below:

```
hourly_delay <- filter(
  summarise(
    group_by(
      filter(
        flights, 
        !is.na(dep_delay)
      ),
      date, hour
    ),
    delay=mean(dep_delay),
    n=n()
  ),
  n>10
)
```

a. What are some problems with this function? 
b. How easy is it to follow the logic of this? 
c. Rewrite this using piping to make it more understandable. 
d. Does your rewritten command give the same results?

#### Part a: 
This doesn't even run because there is no "date" variable. It is extremely chaotic and I hate the arrow instead of equal sign. They need to at least write the arguments of filter on the same line.

#### Part b: 
It is very hard to follow the logic.

#### Part c: 
```{r}
hourly_delay = 
  flights %>%
    unite("date", 1:3, sep = "-" ) %>%
    filter(!is.na(dep_delay)) %>%
    group_by(date, hour) %>%
    summarise(delay = mean(dep_delay), n = n()) %>%
    filter(n > 10)%>%
    arrange(date, hour)
```

#### Part d: 
There weren't results from the previous version so it is impossible to compare. However, if there were an explicit date variable, I think they would give the same results. 

## Creative Exploration Part: 
There is a data set called "weather." I would compare the "wind speed", "precip", and "visib" variables to the dates of the most delayed flights to see if there is any underlying common weather theme. To do this more quantitatively, I would perform a regression and see if there is a statistical relationship. 



